# Reshaping

It's fairly common for datasets from public sources to come in formats that need to be reshaped. The World Development Indicators (WDI) is one such dataset that requires reshaping before we can analyse it. Let's go over the steps to see how we can reshape the WDI dataset.

Let's start by loading the `tidyverse` package first.

```{r message = FALSE}
library(tidyverse)
```

Clear everything to make sure there's nothing leftover in our environment

```{r}
rm(list = ls())
```

We're using a small sample of the WDI dataset here to simplify the tasks. Let's load the dataset and see what it looks like. 

```{r message = FALSE}
wdi <- read_csv("https://raw.githubusercontent.com/SigWeber/data-manipulation-workshop/master/data/wdi.csv", na = "..")

wdi
```

But ideally, we'd like our data to look something like this:

```{r echo = FALSE}
wdi %>%
  filter(Country.Code != "") %>% 
    select(-Series.Code) %>%
  rename(Indicator = `¬Series.Name`) %>% 
  pivot_longer(cols = starts_with("X"),
               names_to = "Year",
               values_to = "Development.Index") %>% 
  mutate(Year = as.numeric(substring(Year, 2, 5)))
```

So, what do we do to achieve this new data format? Note: We want to move away from a **wide** data frame to a **long** data frame!

Step 1: We can see that some country names and codes are blank, so let's get rid of them first

```{r}
wdi %>%
  filter(Country.Code != "") 
```

So far so good. Note that we're not making any changes yet so we can just add one function at a time to the pipeline and check the results. Once we're satisfied with the results we save them to a variable.

Step 2: The dataset contains maternal mortality and health expenditure rates but I really don't like the variable names and what use do I have for the series code? I want to rename and exclude some variables!

```{r}
wdi %>%
  filter(Country.Code != "") %>% 
    select(-Series.Code) %>%
  rename(Indicator = `¬Series.Name`)
```

That looks already but we still have to tackle our main task: how do we get all data entries for different years into one row instead of 3 columns? We want to put all etnries in the columns starting with an X to be below each other. The function to achieve this is *pivot_longer()*. All we have to do is to specify the data and to select the columns we want to reformat. However, we still want to know which value of the world development index was achieved in which year. So, we also tell this to the function: 

```{r }
wdi %>%
  filter(Country.Code != "") %>% 
    select(-Series.Code) %>%
  rename(Indicator = `¬Series.Name`) %>% 
  pivot_longer(cols = starts_with("X"),
               names_to = "Year",
               values_to = "Development.Index")
```

Thats already pretty close. The Year column looks ugly but that's an easy fix. We can use the `substring()` function to take all the characters from position `2` to `5` and assign it back to the `Year` column.

Since this is the last step we might as well assign the results to a new dataset.

```{r }
wdi_long <- wdi %>%
  filter(Country.Code != "") %>% 
    select(-Series.Code) %>%
  rename(Indicator = `¬Series.Name`) %>% 
  pivot_longer(cols = starts_with("X"),
               names_to = "Year",
               values_to = "Development.Index") %>% 
  mutate(Year = as.numeric(substring(Year, 2, 5)))

wdi_long
```

Sometimes, but less often, you will also reshape data to a wide format. That can be achieved with the sibling function *pivot_wider()* in the tidyverse. Check out the online help for this function if you want to use it. 

